---
title: "Differential Abundance Exploration"
author: "Norbert Szala"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
library(ggplot2)
library(forcats)
library(stringr)

source("scripts/functions/00_common_functions.R")
source("scripts/functions/06_differential_abundance_functions.R")

theme_set(theme_minimal(base_size = 12))
set.seed(123)
```

# 1. Aim

This notebook explores differentially abundant taxa and functional pathways between healthy/control and IBD samples. It is a copy of 06_differential_abundance_functions script.

The analysis uses:

- Wilcoxon rank-sum test for each feature,
- Benjamini-Hochberg FDR correction,
- median abundance difference as the main effect direction,
- top-ranked taxa and pathways for visualization.

# 2. Input paths

```{r paths}
input_files <- list(
  taxa = "data/processed/taxa_matrix.csv",
  pathways = "data/processed/pathway_matrix.csv",
  metadata = "data/processed/metadata.csv"
)

output_files <- list(
  taxa_results = "results/tables/differential_taxa.csv",
  pathways_results = "results/tables/differential_pathways.csv",
  top_features = "results/tables/differential_top_features.csv",
  taxa_plot = "results/figures/top_differential_taxa.png",
  pathway_plot = "results/figures/top_differential_pathways.png"
)

params <- list(
  ntop_features = 20
)

walk(
  dirname(unlist(output_files)),
  dir.create,
  recursive = TRUE,
  showWarnings = FALSE
)
```

# 3. Load data

```{r load-data}
taxa_mat <- load_abundance_matrix(input_files$taxa)
pathways_mat <- load_abundance_matrix(input_files$pathways)
metadata <- load_metadata(input_files$metadata)
```

```{r inspect-dimensions}
head(taxa_mat)
head(pathways_mat)
head(metadata)

dim(taxa_mat)
dim(pathways_mat)
dim(metadata)
```

```{r metadata-summary}
metadata %>%
  count(disease_status)
```

# 4. Match samples

```{r match-samples}
taxa_matched <- match_samples(taxa_mat, metadata)
pathways_matched <- match_samples(pathways_mat, metadata)

taxa_mat <- taxa_matched$mat
taxa_metadata <- taxa_matched$metadata

pathways_mat <- pathways_matched$mat
pathways_metadata <- pathways_matched$metadata
```

```{r check-sample-order}
identical(rownames(taxa_mat), taxa_metadata$sample_id)
identical(rownames(pathways_mat), pathways_metadata$sample_id)
```

# 5. Convert matrices to long format

```{r long-format}
taxa_long_df <- matrix_to_long_df(
  mat = taxa_mat,
  feature_set = "taxa"
) %>%
  left_join(
    taxa_metadata %>% select(sample_id, disease_status),
    by = "sample_id"
  )

pathways_long_df <- matrix_to_long_df(
  mat = pathways_mat,
  feature_set = "pathways"
) %>%
  left_join(
    pathways_metadata %>% select(sample_id, disease_status),
    by = "sample_id"
  )
```

```{r inspect-long-tables}
head(taxa_long_df)
head(pathways_long_df)

taxa_long_df %>% count(disease_status)
pathways_long_df %>% count(disease_status)
```

```{r check-missing-status}
sum(is.na(taxa_long_df$disease_status))
sum(is.na(pathways_long_df$disease_status))
```

# 6. Test one feature

Before running the analysis for all features, test one feature manually.

```{r test-one-taxa-feature}
one_taxa_feature <- taxa_long_df %>%
  filter(feature == first(feature))

one_taxa_feature %>%
  count(disease_status)

wilcoxon_test(
  value = one_taxa_feature$value,
  group = one_taxa_feature$disease_status
)
```

# 7. Run differential abundance analysis

```{r run-da}
taxa_results <- run_differential_abundance(taxa_long_df)
pathways_results <- run_differential_abundance(pathways_long_df)

all_results <- bind_rows(taxa_results, pathways_results)

top_features <- get_top_features(
  results_df = all_results,
  ntop = params$ntop_features
)
```

# 8. Inspect results

## Taxa

```{r top-taxa-table}
taxa_results %>%
  select(
    feature,
    median_healthy,
    median_ibd,
    difference_median,
    prevalence_healthy,
    prevalence_ibd,
    p_value,
    p_adj,
    direction
  ) %>%
  arrange(p_adj, desc(abs(difference_median))) %>%
  head(20)
```

## Pathways

```{r top-pathways-table}
pathways_results %>%
  select(
    feature,
    median_healthy,
    median_ibd,
    difference_median,
    prevalence_healthy,
    prevalence_ibd,
    p_value,
    p_adj,
    direction
  ) %>%
  arrange(p_adj, desc(abs(difference_median))) %>%
  head(20)
```

# 9. Number of significant features

```{r significant-counts}
all_results %>%
  mutate(significant = p_adj < 0.05) %>%
  count(feature_set, significant)
```

# 10. Visualize top features

## Top taxa

```{r plot-top-taxa, fig.width=9, fig.height=6}
taxa_plot <- plot_top_features(
  results_df = taxa_results,
  selected_feature_set = "taxa",
  title = paste("Top", params$ntop_features, "differentially abundant taxa"),
  ntop = params$ntop_features
)

taxa_plot
```

## Top pathways

```{r plot-top-pathways, fig.width=9, fig.height=6}
pathways_plot <- plot_top_features(
  results_df = pathways_results,
  selected_feature_set = "pathways",
  title = paste("Top", params$ntop_features, "differentially abundant pathways"),
  ntop = params$ntop_features
)

pathways_plot
```

# 11. Save outputs

```{r save-outputs}
readr::write_csv(taxa_results, output_files$taxa_results)
readr::write_csv(pathways_results, output_files$pathways_results)
readr::write_csv(top_features, output_files$top_features)

ggsave(
  filename = output_files$taxa_plot,
  plot = taxa_plot,
  width = 8,
  height = 6,
  dpi = 300
)

ggsave(
  filename = output_files$pathway_plot,
  plot = pathways_plot,
  width = 8,
  height = 6,
  dpi = 300
)
```


# 12. Session info

```{r session-info}
sessionInfo()
```